### 2017-01-23 doc/manuscript.Rmd
Confirmed manuscript created as expected (minor edits, fixes)

---
bibliography: 
    - /home/luke/Documents/bibliography/bib/bib-db/tagDiabetes.bib
    - /home/luke/Documents/bibliography/bib/bib-db/myrefs.bib
    - /home/luke/Documents/bibliography/bib/bib-db/methods.bib
output: fost::default_manuscript
---

@@

@@

```{r load_data, message=FALSE}
devtools::load_all('.')
ds0yr <- project_data %>% 
    dplyr::filter(VN == 0)
```

@@

Recruitment for the baseline visit of the Prospective Metabolism and Islet Cell Evaluation (PROMISE) 
cohort took place between 2004-2006 from London and Toronto, 
Canada. Individuals were selected to participant if they met the eligibility 
criteria of having one or more risk factors for type 2 diabetes mellitus, 
including obesity, hypertension, family history of diabetes, and/or a history of
gestational diabetes or birth of a macrosomic infant. A total of 736 individuals
attended the baseline visit. Subsequent examinations occurred every three years,
with three examination visits completed to date (2004-2006, 2007-2009, and
2010-2013). Participants are contacted annually by telephone. The current study
sample used data on participants who did not have diabetes at baseline, who
returned for one or more of the follow-up examinations, and who had samples
available for fatty acid measurements (n=477; see the CONSORT diagram in 
`r cite_sf('consort')`; a total of `r calc_n_for_visits()[1]` attended all three
visits). Metabolic characterization, anthropometric
measurements, and questionnaires on lifestyle and sociodemographics are
collected at each examination visit. Research ethics approval was obtained from
Mount Sinai Hospital and the University of Western Ontario, and all participants
provided written informed consent. Data collection methods were standardized
across the 2 centres and research nurses were centrally trained.

@@

TAGFA composition was quantified using stored fasting serum samples from the 
baseline visit, which had been frozen at -70°C for 4-6 years and had not been 
exposed to any freeze-thaw cycles. Serum fatty acids have been documented to be 
stable at these temperatures for up to 10 years [@Matthan2010a]. A known amount 
of {{ confirm with Richard the correct name}} heptadecanoic acid (17:0) was added as an internal
standard prior to extracting total lipids according to the method of Folch 
[@Folch1957a]. Each serum lipid fraction---non-esterified fatty acids (NEFA), 
cholesteryl ester, phospholipid, and TAGFA---was isolated using thin layer 
chromatography. Fatty acid methyl esters were separated and quantified using a 
Varian-430 gas chromatograph (Varian, Lake Forest, CA, USA) equipped with a 
Varian Factor Four capillary column and a flame ionization detector. Fatty acid 
concentrations (nmol/ml) were calculated by proportional comparison of gas
chromatography peak areas to that of the internal standards [@Nishi2014a]. There
were 22 fatty acids measured in the TAGFA fraction. Given that each lipid
fraction differs in biology and given the complexity of the analyses, findings
for other lipid fractions in this cohort are reported separately (see ref
@Johnston2016a for the analysis of the phospholipid and cholesteryl ester
fractions). 
<!--and {{next citation}} for the analysis of the NEFA fraction).-->

@@

Covariates were selected based on the previous literature, from directed acyclic
graph [@Greenland1999a] recommendations (using the DAGitty software
[@Textor2011a], http://dagitty.net/; see `r cite_f('dag_is')` and `r cite_f('dag_bcf')`), and from
QIC. `r cite_t('tab_qic')` shows the covariates compared using QIC.
While the final GEE model selected as best fitting differed between insulin
sensitivity and beta-cell function measures, the differences in QIC values were
less than 10 between many of the models, suggesting similar fit. As such, we selected
the model that had the fewest covariates and that had similar fit between the
outcome measures ({{##}}). The final GEE model was adjusted for {{covariates}}.
After scaling, log-transforming, and exponentiating, the GEE estimates are 
interpreted as an expected percent difference in the outcome variable for every 
standard deviation (SD) increase in the predictor variable given the covariates 
are held constant (including time). Lastly for the GEE models, we tested for an interaction with 
sex, ethnicity, or time by the predictor term for each outcome variable.

@@

`r cite_t('tab_basic')` shows some basic characteristics of the PROMISE cohort. The mean 
follow-up time was `r calc_followup_time()`, where `r calc_pct_full_visits()`%
of participants attended all three visits. There were
`r female$npct` females and `r euro$npct` who had European-ancestry, with a mean
age in years of `r aide::ave_sd(ds0yr$Age)` and a mean BMI
of `r aide::ave_sd(ds0yr$BMI)`. Given the eligibility criteria, the majority of
participants (`r fhd$npct`) had a family history of diabetes.
Between the baseline visit and the 6-year visit in this sample, insulin
sensitivity and beta-cell function measures had a significant median decline
between `r outcome_chg$chg` (p`r outcome_chg$p` from GEE; n=`r outcome_chg$n`). There 
were `r dys$dm` and `r dys$pre` incident cases of diabetes and
pre-diabetes (IFG and IGT), respectively.

`r cite_f('fig_tagfa')` shows the composition of each FA in the TAG fraction (see 
`r cite_t('tab_tagfa')` for a tabular presentation of the values). 
Three TAGFA contributed `r format_rounding(sum(tagfa$pct))`% to the total 
TAG concentration: `r tagfa$c[1]`; `r tagfa$c[2]`; and, `r tagfa$c[3]`.
`r fig_corr` shows a heatmap of the correlation of individual TAGFA with the
outcome variables and several basic characteristics.
As expected, nearly all TAGFA had very strong positive correlations (r=
`r fa_cor$rng$TAG`) with clinically-measured TAG and moderate positive 
correlations with WC (r=`r fa_cor$rng$Waist`). There was also moderate
negative correlations with HDL (r=`r fa_cor$rng$HDL`). For the outcome variables,
the correlations for the insulin sensitivity measures were generally higher 
(HOMA2-%S: r=`r fa_cor$rng['HOMA2-%S']`, ISI: r=`r fa_cor$rng$ISI`) than 
for the beta-cell function measures (all r<0.30).

@@

There were `r nrow(dplyr::filter(gee_unadj_df, p.value <= 0.05))` FDR-corrected 
significant associations in the unadjusted GEE model (see `r cite_f('fig_gee_unadj')`) and 
`r nrow(dplyr::filter(gee_adj_df, p.value <= 0.05))` FDR-corrected
significant associations in the adjusted GEE model as shown in `r cite_f('fig_gee_adj')`. 
The majority of associations with beta-cell function measures were attenuated
after full model adjustment, while nearly all associations with insulin
sensitivity remained significant. {{ confirm }} Subsequent analysis revealed
that the attenuation with beta-cell function was due primarily to adjustment for
{{ waist? ALT? }}.

@@

Both clinically-measured TAG and total TAGFA concentration had very strong
negative associations with all outcome variables. There were no significant
interactions by time, sex, or ethnicity for any of the TAGFA (data not shown).
A tabular presentation of the GEE results is shown in `r cite_t("tab_gee_unadj")` 
for unadjusted models and `r cite_t("tab_gee_adj")` for adjusted models.

@@

The PLS results for the insulin sensitivity (`r cite_f('fig_pls')`) revealed a clustering 
of the fatty acids 14:0, 14:1n-7, 16:0, and 16:1n-7. These TAGFA loaded strongly (meaning
the fatty acids fell between the dashed and solid circles) and negatively on HOMA2-%S
and ISI, suggesting this cluster of TAGFA tracks together with lower insulin
sensitivity. The TAGFA 20:2n-6, 20:5n-3, 22:5n-3, and 22:6n-3 loaded positively
on both insulin sensitivity measures. No other TAGFA loaded strongly, the
results of which remained after running the PLS model on the cross-validated
testing set. The PLS model for insulin sensitivity had good predictive ability, 
determined by correlating the predicted outcome values with the actual values
(HOMA2-%S: `r cor_pls_homa2$r_p`; ISI: `r cor_pls_isi$r_p`).

@@

To our knowledge, no longitudinal study has examined the role of the composition
of the TAGFA fraction on detailed OGTT-derived metabolic measures using advanced
{{another word? appropriate?}} statistical techniques designed to maximally
utilize the data. Two large prospective studies have been published that had
similar research objectives as the present analysis. One was a nested
case-control cohort (n=189 cases and n=189 controls) of the Framingham offspring
study [@Rhee2015a], which found that TAGFA composition with a lower carbon chain
and less double bonds (e.g. 14:0, 16:0) had a higher risk for diabetes after 
12-years while higher carbon chain and more double bond TAGFA had a lower risk 
for diabetes. This pattern of TAGFA with diabetes risk remained the same when 
they examined HOMA-IR cross-sectionally at the baseline visit. The other study 
was a prospective cohort of males in Finland [@Lankinen2015a], of which they had
TAGFA data on 831 participants after 6-years from the baseline visit. This 
cohort has similar outcomes as PROMISE, however, OGTT data was only available at
the 6-year visit. A beta-cell function measure was computed, but only included 
calculations up to 30 minutes of the OGTT. As with our results, they found a 
larger effect size for insulin sensitivity compared to the beta-cell function. 
The specific TAGFA that associated with the metabolic measures were similar 
between studies, particularly with the four DNL fatty acids. A limitation of 
the study was the simpler statistical techniques used to analyze the data, 
which are not able to adequately use the longitudinal and high-dimensional 
dataset. {{ They're article and methods were ... questionable... }} In another 
study of a much smaller (n=16) mostly female group [@Kotronen2009a], the authors
report a positive correlation between total esterified (of which TAG make up the
majority) 16:0, 16:1n-7, and 18:1n-9 with HOMA-IR, similar as our results except
for the 18:1n-9 finding.

@@

Serum TAG are packaged into lipoproteins, particularly VLDL, by the liver. The
sources of fatty acids for TAG production during fasting are stored fatty acids
within the liver, circulating NEFA, and through DNL [@Barrows2006a]. Therefore, the
TAGFA composition reflects not just dietary intake but also stored fatty acids
from the adipose tissue and the DNL capacity of the individual. During fasting, a
large portion of fatty acids bound together as TAG are obtained from circulating
NEFA. A study examining this found that NEFA contributed 60-80% of newly synth
VLDL-TAG during fasting and ~40% during the fed state while dietary fatty acids
contributed ~25% to newly synth VLDL-TAG [@Barrows2006a]. In an analysis of the
serum NEFA fraction in the PROMISE cohort with the OGTT-derived outcomes
(currently unpublished), we found that higher total NEFA but not the specific
composition associated with lower beta-cell function and not with insulin 
resistance. While NEFA may contribute substantially to the TAG fraction, there 
may be something unique about the specific composition of the TAGFA fraction 
that may allow it to enter insulin sensitive tissues more easily than NEFA. It 
may be that VLDL receptors allow fatty acids to enter these tissue more easily
than NEFA bound to albumin, or it may be due to the other influences on TAGFA 
composition such as dietary intake.

@@

The number of components to extract from PLS was determined by 10-fold CV.
Dataset was split into a training set and a testing set to determine the
predictability of the given PLS model. Final results presented use the full dataset.a

>>>>

---
bibliography: 
    - /home/luke/Documents/bibliography/bib/bib-db/tagDiabetes.bib
    - /home/luke/Documents/bibliography/bib/bib-db/myrefs.bib
    - /home/luke/Documents/bibliography/bib/bib-db/methods.bib
csl: /home/luke/Documents/bibliography/styles/dependent/diabetes-care.csl
output: fost::default_manuscript
---

@@

Word count: ~4200

@@

```{r load_data, message=FALSE, echo=FALSE}
devtools::load_all('.')
ds0yr <- project_data %>% 
    dplyr::filter(VN == 0)
```

@@

Recruitment for the baseline visit of the Prospective Metabolism and Islet Cell Evaluation (PROMISE) 
cohort took place between 2004-2006 from London and Toronto, 
Canada. Individuals were selected to participant if they met the eligibility 
criteria of having one or more risk factors for type 2 diabetes mellitus, 
including obesity, hypertension, family history of diabetes, and/or a history of
gestational diabetes or birth of a macrosomic infant. A total of 736 individuals
attended the baseline visit. Subsequent examinations occurred every three years,
with three examination visits completed to date (2004-2006, 2007-2009, and
2010-2013). Participants are contacted annually by telephone. The current study
sample used data on participants who did not have diabetes at baseline, who
returned for one or more of the follow-up examinations, and who had samples
available for fatty acid measurements (n=477; see the CONSORT diagram in 
`r cite_f('consort')`; a total of `r calc_n_for_visits()[1]` attended all three
visits). Metabolic characterization, anthropometric
measurements, and questionnaires on lifestyle and sociodemographics are
collected at each examination visit. Research ethics approval was obtained from
Mount Sinai Hospital and the University of Western Ontario, and all participants
provided written informed consent. Data collection methods were standardized
across the 2 centres and research nurses were centrally trained.

@@

TAGFA composition was quantified using stored fasting serum samples from the 
baseline visit, which had been frozen at -70°C for 4-6 years and had not been 
exposed to any freeze-thaw cycles. Serum fatty acids have been documented to be 
stable at these temperatures for up to 10 years [@Matthan2010a]. A known amount 
of {{confirm with Richard the correct name}} heptadecanoic acid (17:0) was added as an internal
standard prior to extracting total lipids according to the method of Folch 
[@Folch1957a]. Each serum lipid fraction (non-esterified fatty acids (NEFA), 
cholesteryl ester, phospholipid, and TAGFA) was isolated using thin layer 
chromatography. Fatty acid methyl esters were separated and quantified using a 
Varian-430 gas chromatograph (Varian, Lake Forest, CA, USA) equipped with a 
Varian Factor Four capillary column and a flame ionization detector. Fatty acid 
concentrations (nmol/ml) were calculated by proportional comparison of gas
chromatography peak areas to that of the internal standards [@Nishi2014a]. There
were 22 fatty acids measured in the TAGFA fraction. Given that each lipid
fraction differs in biology and given the complexity of the analyses, findings
for other lipid fractions in this cohort are reported separately (see ref
@Johnston2016a for the analysis of the phospholipid and cholesteryl ester
fractions). 
<!--and {{next citation}} for the analysis of the NEFA fraction).-->

@@

Covariates were selected based on the previous literature, from directed acyclic
graph [@Greenland1999a] recommendations (using the DAGitty software
[@Textor2011a], http://dagitty.net/; see `r cite_f('dag_is')` and `r cite_f('dag_bcf')`), and from
QIC. `r cite_t('qic')` shows the covariates compared using QIC.
While the final GEE model selected as best fitting differed between insulin
sensitivity and beta-cell function measures, the differences in QIC values were
less than 10 between many of the models, suggesting similar fit. As such, we selected
the model that had the fewest covariates and that had similar fit between the
outcome measures. The final GEE model was adjusted for time, waist
circumference, baseline age, ethnicity, sex, ALT, MET, and total NEFA.
After scaling, log-transforming, and exponentiating, the GEE estimates are 
interpreted as an expected percent difference in the outcome variable for every 
standard deviation (SD) increase in the predictor variable given the covariates 
are held constant (including time). Lastly for the GEE models, we tested for an
interaction with sex, ethnicity, or time by the predictor term for each outcome
variable.

@@

`r cite_t('basic')` shows some basic characteristics of the PROMISE cohort. The mean 
follow-up time was `r calc_followup_time()`, where `r calc_pct_full_visits()`%
of participants attended all three visits. There were
`r female$npct` females and `r euro$npct` who had European-ancestry, with a mean
age in years of `r aide::ave_sd(ds0yr$Age)` and a mean BMI
of `r aide::ave_sd(ds0yr$BMI)`. Given the eligibility criteria, the majority of
participants (`r fhd$npct`) had a family history of diabetes.
Between the baseline visit and the 6-year visit in this sample, insulin
sensitivity and beta-cell function measures had a significant median decline
between `r outcome_chg$chg` (p`r outcome_chg$p` from GEE; n=`r outcome_chg$n`). There 
were `r dys$dm` and `r dys$pre` incident cases of diabetes and
pre-diabetes (IFG and IGT), respectively.

`r cite_f('tagfa')` shows the composition of each FA in the TAG fraction (see 
`r cite_t('tagfa')` for a tabular presentation of the values). 
Three TAGFA contributed `r format_rounding(sum(tagfa$pct))`% to the total 
TAG concentration: `r tagfa$c[1]`; `r tagfa$c[2]`; and, `r tagfa$c[3]`.
`r cite_f('heatmap')` shows a heatmap of the correlation of individual TAGFA with the
outcome variables and several basic characteristics.
As expected, nearly all TAGFA had very strong positive correlations (r=
`r fa_cor$rng$TAG`) with clinically-measured TAG and moderate positive 
correlations with WC (r=`r fa_cor$rng$Waist`). There was also moderate
negative correlations with HDL (r=`r fa_cor$rng$HDL`). For the outcome variables,
the correlations for the insulin sensitivity measures were generally higher 
(HOMA2-%S: r=`r fa_cor$rng['HOMA2-%S']`, ISI: r=`r fa_cor$rng$ISI`) than 
for the beta-cell function measures (all r<0.30).

@@

There were `r nrow(dplyr::filter(gee_unadj_df, p.value <= 0.05))` FDR-corrected 
significant associations in the unadjusted GEE model (see `r cite_f('gee_unadj')`) and 
`r nrow(dplyr::filter(gee_adj_df, p.value <= 0.05))` FDR-corrected
significant associations in the adjusted GEE model as shown in `r cite_f('gee_adj')`. 
The majority of associations with beta-cell function measures were attenuated
after full model adjustment, while nearly all associations with insulin
sensitivity remained significant. 
<!--{{ confirm }} Subsequent analysis revealed
that the attenuation with beta-cell function was due primarily to adjustment for
{{ waist? ALT? }}.-->

@@

Both clinically-measured TAG and total TAGFA concentration had very strong
negative associations with all outcome variables. There were no significant
interactions by time, sex, or ethnicity for any of the TAGFA (data not shown).
A tabular presentation of the GEE results is shown in `r cite_t("gee_unadj")` 
for unadjusted models and `r cite_t("gee_adj")` for adjusted models.

@@

The PLS results for the insulin sensitivity (`r cite_f('pls')`) revealed a clustering 
of the fatty acids 14:0, 14:1n-7, 16:0, and 16:1n-7. These TAGFA loaded strongly (meaning
the fatty acids fell between the dashed and solid circles) and negatively on HOMA2-%S
and ISI, suggesting this cluster of TAGFA tracks together with lower insulin
sensitivity. The TAGFA 20:2n-6, 20:5n-3, 22:5n-3, and 22:6n-3 loaded positively
on both insulin sensitivity measures. No other TAGFA loaded strongly, the
results of which remained after running the PLS model on the cross-validated
testing set. The PLS model for insulin sensitivity had good predictive ability, 
determined by correlating the predicted outcome values with the actual values
(HOMA2-%S: `r cor_pls_homa2$r_p`; ISI: `r cor_pls_isi$r_p`).

@@

To our knowledge, no longitudinal study has examined the role of the composition
of the TAGFA fraction on detailed OGTT-derived metabolic measures using advanced
{{another word? appropriate?}} statistical techniques designed to maximally
utilize the data. Two large prospective studies have been published that had
similar research objectives as the present analysis. One was a nested
case-control cohort (n=189 cases and n=189 controls) of the Framingham offspring
study [@Rhee2011a], which found that TAGFA composition with a lower carbon chain
and less double bonds (e.g. 14:0, 16:0) had a higher risk for diabetes after 
12-years while higher carbon chain and more double bond TAGFA had a lower risk 
for diabetes. This pattern of TAGFA with diabetes risk remained the same when 
they examined HOMA-IR cross-sectionally at the baseline visit. The other study 
was a prospective cohort of males in Finland [@Lankinen2015a], of which they had
TAGFA data on 831 participants after 6-years from the baseline visit. This 
cohort has similar outcomes as PROMISE, however, OGTT data was only available at
the 6-year visit. A beta-cell function measure was computed, but only included 
calculations up to 30 minutes of the OGTT. As with our results, they found a 
larger effect size for insulin sensitivity compared to the beta-cell function. 
The specific TAGFA that associated with the metabolic measures were similar 
between studies, particularly with the four DNL fatty acids. A limitation of 
the study was the simpler statistical techniques used to analyze the data, 
which are not able to adequately use the longitudinal and high-dimensional 
dataset. {{ Their article and methods were ... questionable... }} In another 
study of a much smaller (n=16) mostly female group [@Kotronen2009a], the authors
report a positive correlation between total esterified (of which TAG make up the
majority) 16:0, 16:1n-7, and 18:1n-9 with HOMA-IR, similar as our results except
for the 18:1n-9 finding.

@@

Serum TAG are packaged into lipoproteins, particularly VLDL, by the liver. The
sources of fatty acids for TAG production during fasting are stored fatty acids
within the liver, circulating NEFA, and through DNL [@Barrows2006a]. Therefore, the
TAGFA composition reflects not just dietary intake but also stored fatty acids
from the adipose tissue and the DNL capacity of the individual. During fasting, a
large portion of fatty acids bound together as TAG are obtained from circulating
NEFA. A study examining this found that NEFA contributed 60-80% of newly synthesized
VLDL-TAG during fasting and ~40% during the fed state while dietary fatty acids
contributed ~25% to newly synthesized VLDL-TAG [@Barrows2006a]. In an analysis of the
serum NEFA fraction in the PROMISE cohort with the OGTT-derived outcomes
(currently unpublished), we found that higher total NEFA but not the specific
composition associated with lower beta-cell function and not with insulin 
resistance. While NEFA may contribute substantially to the TAG fraction, there 
may be something unique about the specific composition of the TAGFA fraction 
that may allow it to enter insulin sensitive tissues more easily than NEFA. It 
may be that VLDL receptors allow fatty acids to enter these tissue more easily
than NEFA bound to albumin, or it may be due to the other influences on TAGFA 
composition such as dietary intake.

@@

<!--
The number of components to extract from PLS was determined by 10-fold CV.
Dataset was split into a training set and a testing set to determine the
predictability of the given PLS model. Final results presented use the full dataset.a
-->
